Individual and Differential Harm in Redistricting

### 2025-08-25 paper/harm.qmd
push

The proposed framework is not without important limitations, however, which present opportunities for future study.
The flexibility of choosing a model for counterfactual plans and hypothetical election outcomes also creates possibly large sensitivity to those choices.
<!-- Fundamentally, harm and differential harm are unable to measure plans against an absolute benchmark, the way that other measures such as partisan symmetry are able to do. -->
Simple election models are computationally tractable and require minimal data, but grossly simplify reality.
More realistic models can add a significant burden to the calculation of our harm measures.
And while modern simulation methods are a valuable starting point for defining counterfactual plans, these approaches involve analyst discretion and interpretation of often-murky legal standards.
Incorrect or incomplete interpretation of these standards can negatively influence the validity of the counterfactual plans.

@@

Even when the choice of constraints maps onto legal requirements, it is important to ensure that the counterfactual plans are similar to the enacted plan on discretionary criteria as well.
For example, a mapdrawer may weakly respect something like compactness or respecting county boundaries without being required to do so.
In this case, if these optional constraints are not matched in the counterfactual, it will be unclear whether any estimated harm is due to *actual* biases in a plan or the mismatch in criteria.
As such, it is important to not just ensure that any counterfactual samples use methods with known sampling properties, but also that the sampled plans are similar to the enacted plan on these discretionary criteria.

@@

A quite different approach to measuring partisan gerrymandering is the partisan dislocation [@defoeubarodd21].
This measure uses a $k$-nearest-neighbor approach, where $k$ is the size of a district, to identify when a voter's district is politically different from their assigned district.
This can be repeated for  voter in a state or region, demonstrating at a local level when people been placed in a district which is atypical for who their nearest neighbors are.
When aggregated to the precinct-level, this can produce local measures of partisan gerrymandering, similar to our proposed differential partisan harm measure.

@@

The partisan dislocation measure [@defoeubarodd21] shows that the Democratic plan is typical ($p=0.24$) while the Republican plan is an outlier ($p=0.01$).
We use the vote-weighted measure of the average dislocation to produce a statewide summary.
We find that the sign is incorrect for the Democratic plan, which has a positive sign.
As the sign is positive for the simulations as well, this difference may be related to the fact that this measure uses the $k$ nearest neighbors, which is not uniform over the voters in the state.
Implicitly, voters in dense, urban areas will be counted at a higher rate than those in rural areas.

Note that the Ranked Marginal Deviation [@herschlag2020quantifying] is not a signed measure.
All scores are at least zero, with low scores indicating that a plan is similar to the simulated plans.
In this analysis, the Democratic plan sits within the range of the sampled plans, while the Republican plan is an outlier ($p=0.0002$).

>>>>

The proposed framework is not without important limitations, however, which present opportunities for future study.
The flexibility of choosing a model for counterfactual plans and hypothetical election outcomes also creates possibly large sensitivity to those choices.
<!-- Fundamentally, harm and differential harm are unable to measure plans against an absolute benchmark, the way that other measures such as partisan symmetry are able to do. -->
Simple election models are computationally tractable and require minimal data, but grossly simplify reality.
More realistic models can add a significant burden to the calculation of our harm measures.
And while modern simulation methods are a valuable starting point for defining counterfactual plans, these approaches involve analyst discretion and interpretation of often-murky legal standards, and face their own computational challenges.
Incorrect or incomplete interpretation of these standards can negatively influence the validity of the counterfactual plans.

@@

Even when the choice of constraints maps onto legal requirements, it is important to ensure that the counterfactual plans are similar to the enacted plan on discretionary criteria as well.
For example, a mapdrawer may target a particular compactness metric, or draw districts that respect county boundaries, without being required to do so.
In this case, if these optional constraints are not matched in the counterfactual, it will be unclear whether any estimated harm is due to *actual* biases in a plan or the mismatch in criteria.
This includes the use of different compactness metrics in simulations than those used by the mapdrawer.
As such, it is important to not just ensure that any counterfactual samples use methods with known sampling properties, but also that the sampled plans are similar to the enacted plan on these discretionary criteria.

@@

A quite different approach to measuring partisan gerrymandering from symmetry-based methods is the partisan dislocation [@defoeubarodd21].
This measure is based on a $k$-nearest-neighbor comparison, where $k$ is the size of a district, to identify when a voter's district is politically different from their assigned district.
This can be repeated for  voter in a state or region, demonstrating at a local level when people been placed in a district which is atypical for who their nearest neighbors are.
When aggregated to the precinct-level or beyond, this can produce local or statewide measures of partisan gerrymandering, similar to our proposed differential partisan harm measure.

However, unlike the counterfactual _plan_ that is used in the harm framework, the comparison $k$-nearest-neighbors do not need to constitute a valid district.
Moreover, harm considers the effect on election _outcomes_, while partisan dislocation measures changes in _vote share_.
In our framework, if a Democrat wins in both the enacted and counterfactual plans, but by a smaller margin, there can be no harm.
Partisan dislocation, in contrast, would measure the change in margin.
This reflects the different quantities of interest: harm to individual voters, rather than changes in partisan configurations.

@@

The partisan dislocation measure [@defoeubarodd21] shows that the Democratic plan is typical ($p=0.24$) while the Republican plan is an outlier ($p=0.01$).
We average each voter's dislocation to produce the statewide summary value shown in the figure.
We find that the sign is incorrect for the Democratic plan, which has positive dislocation (indicating an average shift towards Republicans).
Notably, the sign is positive for the simulations as well, which suggests that such a statewide average of partisan dislocation may not reliably measure partisan bias.
It is difficult to know exactly what causes this positive skew, but we suspect it is due to the interaction of the $k$-nearest neighbors comparison and the state's boundaries and varying population density.

Note that the Ranked Marginal Deviation [@herschlag2020quantifying] is not a signed measure.
All scores are at least zero, with low scores indicating that a plan is similar to the simulated plans.
In this analysis, the Democratic plan sits within the range of the sampled plans, while the Republican plan is an outlier ($p=0.0002$).
This matches the distribution over Democratic seats, as well as average harm.
